---
title: "Do anti-depressants prescription rise in winter months and/or in household with lack of central heating?"
author: "B209732"
date: "2025-11-14"
output:
  html_document:
    code_folding: hide
  
---

Seasonal affective disorder (SAD) is a type of depression also known as the 'winter depression' because it's symptoms are more prevalent and severe during winter months. According to the NHS website, one treatment for SAD is the prescription of anti-depressants (ADs) such as selective serotonin re-uptake inhibitors (SSRIs) (https://www.nhs.uk/mental-health/conditions/seasonal-affective-disorder-sad/overview/ - accessed November 13th, 2025)

This report is going to investigate the link between AD prescriptions from January to June 2025, to see whether prescriptions were higher during winter months (January and February) compared to spring (March to June). Prescription data will be subdivided into Health Boards to see whether monthly trends are observed across different parts of Scotland and to see the links between number of households without central heating in each health board and and AD prescription.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
#Data packages used
library(tidyverse)
library(janitor)
library(gt)
library(here)
library(sf)
library(dbplyr)
```
Here we are opening the prescription data (available here:), Health Board codes and names (available here:), housing data (from census 2022) and population data (from census 2022)
```{r}
#Open data - January 2025 to June 2025 
data_jan_to_june2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

#Health Boards names and codes
HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()

#Central heating data
central_heating <- read_csv(here("data", "central_heating.csv"), skip = 10) %>% #skip the first 10 rows
  clean_names() %>% 
rename(hb_name = "health_board_area_2019",
      n_households = "count") %>% 
  select(hb_name, central_heating_in_household, n_households) %>% 
  filter(str_detect(central_heating_in_household, "All occupied households|No central heating")) %>% 
   mutate(hb_name = paste("NHS", hb_name))

population_data <- read_csv(here("data/UV102a_age_health_board_census.csv"), skip = 10) %>%
  # Rename the last column to avoid the messy name in column 6 and to match column names with the prescription dataset
  rename(Spare = "...6",
         hb_name = "Health Board Area 2019",
         hb_population = Count) %>%  # filter the data so that we get the population of the entire health board
  filter(Age == "All people" & Sex == "All people") %>% 
  # select only the relevant columns
  select(hb_name, hb_population) %>% 
  # change health board names so they match the prescription data
  mutate(hb_name = paste("NHS", hb_name))

```
In this report we will be focusing on SSRIs. The SSRIs selected will be sertraline, citalopram, escitalopram, fluoxetine and paroxetine of all formulations.
```{r}
#filter data 
data_jan_to_june2025 <- data_jan_to_june2025 %>% 
  filter(str_detect(bnf_item_description, "SERTRALINE|CITALOPRAM|ESCITALOPRAM|FLUOXETINE|PAROXETINE")) %>%  full_join(HB_lookup, by = c("hbt" = "hb")) %>% #joined the Health Board data with names and codes
   select(hb_name, bnf_item_description, paid_quantity, paid_date_month) %>% 
  mutate(Months = str_replace_all(paid_date_month ,c('202501'='January', '202502'='February', '202503'='March', '202504'='April', '202505'='May', '202506'='June'))) %>%  #replaced the numerical date with a character one
  full_join(population_data) %>% 
  mutate(AD_per_1k = (paid_quantity/hb_population)*1000)
```

Table 1 and Figure 1:
```{r}
data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k, hb_population) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  summarise(total_per_HB = sum(AD_per_1k, na.rm = TRUE)) %>% 
    gt() %>% 
  cols_label(total_per_HB = "Total AD quantity",
             hb_name = "Health Board") %>% 
  tab_header(title = "Total anti-depressants prescrption per 1k of population",
             subtitle = "Data from NHS Borders and NHS Fife")

#total per HB
data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  ggplot(aes(x = hb_name, y = AD_per_1k)) +
  geom_col() +
  labs(y = "Paid quantity per 1k population", 
       x = "Health Boards",
       title = "Anti-depressants prescription in each Health Board") +
  theme_light()+
  coord_flip()

 
```
Show the current distribution of AD per 1k of population

THIS IS A PLOT THAT SHOWS THE DATA FROM THE TABLE ABOVE
```{r}

data_jan_to_june2025 %>% 
  select(paid_quantity, Months) %>% 
  drop_na() %>% 
  group_by(Months) %>% 
  summarise(total_per_month = sum(paid_quantity, na.rm = TRUE)) %>% 
  gt() %>% 
  cols_label(total_per_month = "Total AD quantity per month") %>% 
  tab_header(title = "Total anti-depressants prescription",
             subtitle = "Quantity per month")

data_jan_to_june2025 %>% 
  select(paid_quantity, Months) %>% 
  drop_na() %>% 
  group_by(Months) %>% 
  ggplot(aes(x = fct_inorder(Months), y = paid_quantity)) + #to make sure ggplot doesn't order months alphabetically
  geom_col() +
  labs(y = "Paid quantity",
       x = "Months",
       title = "Anti-depressants per months") +
  theme_classic()

```
This data shows that there is not a trend in which AD in prescribed more in winter months, it looks like there is a wave of prescription

TRY TO DIVIDE DATA FOR EACH MONTH
```{r}

data_jan_to_june2025 %>% 
  select(AD_per_1k, Months, hb_name) %>% 
  drop_na() %>% 
  group_by(hb_name) %>% 
   ggplot(aes(x = hb_name, y = AD_per_1k)) + 
  geom_col() +
  labs(y = "Paid quantity",
       x = "Health Boards",
       title = "Trends of anti-depressants prescription per Health Board, from January to June") +
  theme_bw() +
  coord_flip() +
  facet_grid(cols = vars(fct_inorder(Months)))

```


