---
title: "Is there a link between anti-depressants (ADs) prescription and lack of central heating?"
author: "B209732"
date: "2025-12-01"
output:
  html_document:
    code_folding: hide
  
---

A study by Clair and Baker (2022) has reported that living in cold homes could have detrimental effects on mental health. Their study, carried out in the UK, has found an increase in mental distress in participants who were unable to keep their house warm. The mental distress experienced by the participants in the study could be attributed to reduced feelings of autonomy, effects on sleep and as a result of feeling cold all the time. The latter is of particular concern in winter months (clair and Baker, 2022). 

Winter is a time in which some people may experience seasonal affective disorder (SAD), as a result of being exposed to less hours of light. This is a type of depression and could be exacerbated by factors such as living in colder houses, such as those lacking central heating.  (https://www.mind.org.uk/information-support/types-of-mental-health-problems/seasonal-affective-disorder-sad/understanding-your-experiences/ - accessed Novermber 3oth, 2025). By looking at the data available in the Scottish Census, a number of households across Scotland currently lack central heating.

This report will investigate the link between lack of central heating and SAD/mental distress by looking at antidepressants (ADs) prescription across all the Health Boards (HBs). This is because ADs are commonly prescribed as a treatment for SAD and mental distress such as increased anxiety. Data across HBs will be used to identify patterns across Scotland (https://www.nhs.uk/mental-health/conditions/seasonal-affective-disorder-sad/overview/ - accessed November 13th, 2025). 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
#Data packages used
library(tidyverse)
library(janitor)
library(gt)
library(here)
library(dplyr)
library(ggplot2)
```
The data used in this report can be found in the Public Health Scotland data sets and Scottish Census 2022:
* Prescription data: https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1
* Health Boards (HB) names and codes: https://www.opendata.nhs.scot/dataset/geography-codes-and-labels/resource/652ff726-e676-4a20-abda-435b98dd7bdc
* housing data: https://www.scotlandscensus.gov.uk/webapi/jsf/tableView/tableView.xhtml
* population data: https://www.scotlandscensus.gov.uk/webapi/jsf/tableView/tableView.xhtml
```{r}
#Open data - January 2025 to June 2025 
data_jan_to_june2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

#Health Boards names and codes
HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()

#Central heating data
central_heating <- read_csv(here("data", "central_heating.csv"), skip = 10) %>% #skip the first 10 rows
  clean_names() %>% 
rename(hb_name = "health_board_area_2019",
      n_households = "count") %>% 
  select(hb_name, central_heating_in_household, n_households) %>% 
  filter(str_detect(central_heating_in_household, "All occupied households|No central heating")) %>% 
   mutate(hb_name = paste("NHS", hb_name))

population_data <- read_csv(here("data/UV102a_age_health_board_census.csv"), skip = 10) %>%
  # Rename the last column to avoid the messy name in column 6 and to match column names with the prescription dataset
  rename(Spare = "...6",
         hb_name = "Health Board Area 2019",
         hb_population = Count) %>%  # filter the data so that we get the population of the entire health board
  filter(Age == "All people" & Sex == "All people") %>% 
  # select only the relevant columns
  select(hb_name, hb_population) %>% 
  # change health board names so they match the prescription data
  mutate(hb_name = paste("NHS", hb_name))

```

For this report, the focus will be on the prescription of selective serotonin re-uptake inhibitors such as sertraline, citalopram, escitalopram, fluoxetine and paroxetine, of all formulations (https://www.nhs.uk/medicines/antidepressants/ - accessed November 30th 2025).

```{r}
#filter SSRIs 
data_jan_to_june2025 <- data_jan_to_june2025 %>% 
  filter(str_detect(bnf_item_description, "SERTRALINE|CITALOPRAM|ESCITALOPRAM|FLUOXETINE|PAROXETINE")) %>%
  full_join(HB_lookup, by = c("hbt" = "hb")) %>% #joined the Health Board data with names and codes
  mutate(Months = str_replace_all(paid_date_month ,c('202501'='January', '202502'='February', '202503'='March', '202504'='April', '202505'='May', '202506'='June'))) %>%  #replaced the numerical date with a character one
  full_join(population_data) %>% #joined the population data
  mutate(AD_per_1k = (paid_quantity/hb_population)*1000) #Since each Health Board has different populations, the AD prescription per 1000 people will be used for comparisons
```

We will start the investigation by looking at the recent distribution of ADs across Health Boards in Scotland

**Table 1: Distribution of AD/1k population across HBs in Scotland**
```{r}
data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k, hb_population) %>%
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  summarise(total_per_HB = sum(AD_per_1k, na.rm = TRUE)) %>% 
    gt() %>% 
  cols_label(total_per_HB = "Total AD quantity",
             hb_name = "Health Board") %>% 
  tab_header(title = "Total anti-depressants prescrption per 1k of population",
             subtitle = "Data from NHS Borders and NHS Fife")

```

Below, we can see a graphical representation of the data presented in the table above.

**Figure 1: Distribution of AD/1k population across HBs in Scotland**
```{r}
#total per HB
data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  ggplot(aes(x = hb_name, y = AD_per_1k)) +
  geom_col() +
  labs(y = "Paid quantity per 1k population", 
       x = "Health Boards",
       title = "Anti-depressants prescription in each Health Board") +
  theme_light()+
  coord_flip()

```

Table and and Figure 1 show the recent distribution of ADs across Scotland. As we can see AD prescription is fairly constant across Health Boards, with some regions of Scotland having slightly higher prescriptions compared to others.

Now we are going to look at the total quantity of AD prescription for each month, to see whether prescription changes according to seasons.

**Table 2: AD prescription per month (January to June 2025)**
```{r}

#change how to calculate total AD we can't add up percentages
data_jan_to_june2025 %>% 
  select(paid_quantity, Months) %>% 
  drop_na() %>% 
  group_by(Months) %>% 
  summarise(total_per_month = sum(paid_quantity, na.rm = TRUE)) %>% 
  gt() %>% 
  cols_label(total_per_month = "Total AD quantity per month") %>% 
  tab_header(title = "Total anti-depressants prescription",
             subtitle = "Quantity per month")
```

**Figure 2: AD prescription per month (January to June 2025)**
```{r}
data_jan_to_june2025 %>% 
  select(paid_quantity, Months) %>% 
  drop_na() %>% 
  group_by(Months) %>% 
  ggplot(aes(x = fct_inorder(Months), y = paid_quantity)) + #to make sure ggplot doesn't order months alphabetically
  geom_col() +
  labs(y = "Paid quantity",
       x = "Months",
       title = "Anti-depressants per months") +
  theme_classic()

```

Table 2 and Figure 2 show that there doesn't seem to be a trend between AD prescription and seasons but rather that a month with high prescription usually follows two months with lower prescriptions. This suggests that colder temperatures don't seem to be associated with increased SAD/mental distress in people living in Scotland.   

Now we are going to look at whether Health Boards trends are stable across months, to identify whether some parts of Scotland are affected by seasons more than others. 

**Figure 3: Comparison of AD prescription per Health Boards, from January to June 2025**
```{r, fig.width=10,fig.height=8} 

data_jan_to_june2025 %>% 
  select(AD_per_1k, Months, hb_name) %>% 
  drop_na() %>% 
  group_by(hb_name) %>% 
   ggplot(aes(x = hb_name, y = AD_per_1k)) + 
  geom_col() +
  labs(y = "Paid quantity",
       x = "Health Boards",
       title = "Trends of anti-depressants prescription per Health Board, from January to June") +
  theme_bw() +
  coord_flip() +
  facet_grid(cols = vars(fct_inorder(Months)))

```
From Figure 3 we can see that AD prescription for each Health Board did not change across months, but maintained a pattern. Therefore, there are not some parts of Scotland that are more affected by SAD/mental distress compared to others. 

Now we are going to look at whether lack of central heating is correlated with higher prescription of ADs. 

```{r}
central_heating_per_HB <- data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k, hb_population) %>%
  drop_na() %>% 
  group_by(hb_name, hb_population) %>% 
  reframe(total_AD_per_HB_per_1k = sum(AD_per_1k, na.rm = TRUE)) %>% #used reframe as summarise did not allow for result to be attributed to more than one row?
  full_join(central_heating) %>% 
  filter(central_heating_in_household == "No central heating") %>% #we wre only interested at houses lacking central heating
  mutate(household_per_1k = (n_households/hb_population)*1000, people_per_household = (hb_population/n_households))
#since each Health Board has different populations, each one will also have a different number of households, therefore we will at number of households per 1k population

central_heating_per_HB %>% 
  summarise(average_n_per_household = mean(people_per_household))#here we estimated how many people usually live in each household - 2.14 people

```

**Table 3: Number of households lackign central heating/1k population across Health Boards** 
```{r}
central_heating_per_HB %>% 
  select(household_per_1k, hb_name, central_heating_in_household) %>% 
  group_by(hb_name) %>% 
  summarise(total_no_heating_per_HB = sum(household_per_1k, na.rm = TRUE)) %>% 
    gt() %>% 
  cols_label(total_no_heating_per_HB = "Total number of households without central heating", hb_name = "Health Boards") %>% 
  tab_header(title = "Total number of households without central heating per 1k population",
             subtitle = "Data from NHS Borders and NHS Fife")

```

**Figure 4: Number of households lackign central heating/1k population across Health Boards** 
```{r}
central_heating_per_HB %>% 
  select(household_per_1k, hb_name, central_heating_in_household) %>% 
  filter(central_heating_in_household == "No central heating") %>% 
  group_by(hb_name) %>% 
  summarise(total_no_heating_per_HB = sum(household_per_1k, na.rm = TRUE)) %>% 
  ggplot(aes(x = hb_name, y = total_no_heating_per_HB)) + 
  geom_col() +
  labs(y = "Households/1k without central heating",
       x = "Health Boards",
       title = "Households/1k in each Health Board") +
  theme_classic()+
  coord_flip()
```

From table 3 and Figure 4 we can see that some Health Boards have more houses lacking central heating compared to others, with the highest number attributed to NHS Orkney.

_Will AD prescription be higher in those Health Boards?_
If so, then there might be a correlation between cold houses and AD prescription.

