---
title: "Do anti-depressants prescription rise in winter months and/or in household with lack of central heating?"
author: "B209732"
date: "2025-11-14"
output:
  html_document:
    code_folding: hide
    latex_engine: xelatex
mainfont: "Times New Roman"
  
---

Seasonal affective disorder (SAD) is a type of depression also known as the 'winter depression' because it's symptoms are more prevalent and severe during winter months. According to the NHS website, one treatment for SAD is the prescription of anti-depressants (ADs) such as selective serotonin re-uptake inhibitors (SSRIs) (https://www.nhs.uk/mental-health/conditions/seasonal-affective-disorder-sad/overview/ - accessed November 13th, 2025)

This report is going to investigate the link between AD prescriptions from January to June 2025, to see whether prescriptions were higher during winter months (January and February) compared to spring (March to June). Prescription data will be subdivided into Health Boards to see whether prescriptions across Health Boards varies or stays stable depending on the months. The links between number of households without central heating in each health board and and AD prescription will also be investigated. A study has suggested that people living in colder households are more likely to develop mental distress even when they didn't have any (reference)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r}
#Data packages used
library(tidyverse)
library(janitor)
library(gt)
library(here)
library(dplyr)
library(ggplot2)
```
Here we will be using prescription data (available here:), Health Board codes and names (available here:), housing data (from census 2022) and population data (from census 2022)
```{r}
#Open data - January 2025 to June 2025 
data_jan_to_june2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

#Health Boards names and codes
HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()

#Central heating data
central_heating <- read_csv(here("data", "central_heating.csv"), skip = 10) %>% #skip the first 10 rows
  clean_names() %>% 
rename(hb_name = "health_board_area_2019",
      n_households = "count") %>% 
  select(hb_name, central_heating_in_household, n_households) %>% 
  filter(str_detect(central_heating_in_household, "All occupied households|No central heating")) %>% 
   mutate(hb_name = paste("NHS", hb_name))

population_data <- read_csv(here("data/UV102a_age_health_board_census.csv"), skip = 10) %>%
  # Rename the last column to avoid the messy name in column 6 and to match column names with the prescription dataset
  rename(Spare = "...6",
         hb_name = "Health Board Area 2019",
         hb_population = Count) %>%  # filter the data so that we get the population of the entire health board
  filter(Age == "All people" & Sex == "All people") %>% 
  # select only the relevant columns
  select(hb_name, hb_population) %>% 
  # change health board names so they match the prescription data
  mutate(hb_name = paste("NHS", hb_name))

```
In this report we will be focusing on SSRIs. The SSRIs selected will be sertraline, citalopram, escitalopram, fluoxetine and paroxetine of all formulations. Since each Health Board has different populations, the AD prescription per 1000 people will be used.
```{r}
#filter data 
data_jan_to_june2025 <- data_jan_to_june2025 %>% 
  filter(str_detect(bnf_item_description, "SERTRALINE|CITALOPRAM|ESCITALOPRAM|FLUOXETINE|PAROXETINE")) %>%
  full_join(HB_lookup, by = c("hbt" = "hb")) %>% #joined the Health Board data with names and codes
  mutate(Months = str_replace_all(paid_date_month ,c('202501'='January', '202502'='February', '202503'='March', '202504'='April', '202505'='May', '202506'='June'))) %>%  #replaced the numerical date with a character one
  full_join(population_data) %>% 
  mutate(AD_per_1k = (paid_quantity/hb_population)*1000) 
```

Table 1 and Figure 1:
```{r}
data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k, hb_population) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  summarise(total_per_HB = sum(AD_per_1k, na.rm = TRUE)) %>% 
    gt() %>% 
  cols_label(total_per_HB = "Total AD quantity",
             hb_name = "Health Board") %>% 
  tab_header(title = "Total anti-depressants prescrption per 1k of population",
             subtitle = "Data from NHS Borders and NHS Fife")

#total per HB
data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  ggplot(aes(x = hb_name, y = AD_per_1k)) +
  geom_col() +
  labs(y = "Paid quantity per 1k population", 
       x = "Health Boards",
       title = "Anti-depressants prescription in each Health Board") +
  theme_light()+
  coord_flip()

```
Table and and Figure 1 show the distribution of AD/1k population across Health Boards. As it can be seen, AD prescription is fairly constant across Health Boards, with some regions of Scotland having slighly lower prescription compared to others.

Now we are going to look at the total quantity of AD prescription for each month, to see whether prescription changes according to seasons.
```{r}

data_jan_to_june2025 %>% 
  select(paid_quantity, Months) %>% 
  drop_na() %>% 
  group_by(Months) %>% 
  summarise(total_per_month = sum(paid_quantity, na.rm = TRUE)) %>% 
  gt() %>% 
  cols_label(total_per_month = "Total AD quantity per month") %>% 
  tab_header(title = "Total anti-depressants prescription",
             subtitle = "Quantity per month")

data_jan_to_june2025 %>% 
  select(paid_quantity, Months) %>% 
  drop_na() %>% 
  group_by(Months) %>% 
  ggplot(aes(x = fct_inorder(Months), y = paid_quantity)) + #to make sure ggplot doesn't order months alphabetically
  geom_col() +
  labs(y = "Paid quantity",
       x = "Months",
       title = "Anti-depressants per months") +
  theme_classic()

```
Table 2 and Figure 2 show that there doesn't seem to be a trend between AD prescription and seasons. However, it seems that a month with high prescription usually seems to follow two months with lower prescriptions, almost like a wave.  

Now we are going to look at whether monthly trends differ or are stable acros Health Boards, to see whether some Health Boards are affected by seasons more than others. 
```{r, fig.width=10,fig.height=8}

data_jan_to_june2025 %>% 
  select(AD_per_1k, Months, hb_name) %>% 
  drop_na() %>% 
  group_by(hb_name) %>% 
   ggplot(aes(x = hb_name, y = AD_per_1k)) + 
  geom_col() +
  labs(y = "Paid quantity",
       x = "Health Boards",
       title = "Trends of anti-depressants prescription per Health Board, from January to June") +
  theme_bw() +
  coord_flip() +
  facet_grid(cols = vars(fct_inorder(Months)))

```
From here we can see that AD prescription for each Health Board did not change across months, but maintained a pattern. 
Now we can conclude that seasonsal changes doesn't affect AD prescription in Scotland.

Now we are going to see whether lack of central heating is correlated with higher prescription of ADs. Since each health board has different populations, they will also have different number of households. Therefore we are going to look at households per 1000 people. Limitations is that we don't know how big these households would be, therefore how many people are in each household (could calcualte how many in each household by diving population by n_household), then household per 1k population.
```{r}
jan_to_june2025_summary <- data_jan_to_june2025 %>% 
  select(hb_name, AD_per_1k, hb_population) %>% 
  drop_na() %>% 
  group_by(hb_name, hb_population) %>% 
  reframe(total_AD_per_HB_per_1k = sum(AD_per_1k, na.rm = TRUE)) %>% 
  full_join(central_heating) %>% 
  mutate(household_per_1k = (n_households/hb_population)*1000, people_per_household = (hb_population/n_households))

jan_to_june2025_summary %>% 
  filter(central_heating_in_household == "All occupied households") %>% 
  summarise(average_n_per_household = mean(people_per_household))

```
By only looking at the data available for all households, we can see that roughly in all Health Boards, 2.14 people live in each household. 
```{r}
jan_to_june2025_summary %>% 
  select(household_per_1k, hb_name, central_heating_in_household) %>% 
  filter(central_heating_in_household == "No central heating") %>% 
  group_by(hb_name) %>% 
  summarise(total_no_heating_per_HB = sum(household_per_1k, na.rm = TRUE)) %>% 
    gt() %>% 
  cols_label(total_no_heating_per_HB = "Total number of households without central heating", hb_name = "Health Boards") %>% 
  tab_header(title = "Total number of households without central heating per 1k population",
             subtitle = "Data from NHS Borders and NHS Fife")

jan_to_june2025_summary %>% 
  select(household_per_1k, hb_name, central_heating_in_household) %>% 
  filter(central_heating_in_household == "No central heating") %>% 
  group_by(hb_name) %>% 
  summarise(total_no_heating_per_HB = sum(household_per_1k, na.rm = TRUE)) %>% 
  ggplot(aes(x = hb_name, y = total_no_heating_per_HB)) + 
  geom_col() +
  labs(y = "Households/1k without central heating",
       x = "Health Boards",
       title = "Households/1k in each Health Board") +
  theme_classic()+
  coord_flip()
```
Here we can see that some Health Boards have more houses without central heating compared to others, with the highest number attributed to NHS Orkney.

_Will AD prescription be higher in those Health Boards?_
If so, then there might be a correlation between cold houses and AD prescription.

```{r}
jan_to_june2025_summary_plot <- jan_to_june2025_summary %>% 
  filter(central_heating_in_household == "No central heating") %>% 
  ggplot(aes(fill = household_per_1k, y = total_AD_per_HB_per_1k, x = hb_name))+
    geom_bar(position = "fill", stat = "identity")+
    coord_flip()
```

