---
title: "Do anti-depressants prescription rise in winter months and/or in the deemest regions of Scotland?"
author: "B209732"
output: html_document
date: "2025-11-14"
---

Seasonal affective disorder (SAD) is a type of depression also known as the 'winter depression' because it's symptoms are more prevalent and severe during winter months. According to the NHS website, one treatment for SAD is the prescription of anti-depressants (ADs) such as selective serotonin re-uptake inhibitors (SSRIs) (https://www.nhs.uk/mental-health/conditions/seasonal-affective-disorder-sad/overview/ - accessed November 13th, 2025)

This report is going to investigate the link between AD prescriptions from January to June 2025, to see whether prescriptions were higher during winter months (January and February) compared to spring (March to June). Prescription data will be subdivided into Health Boards to investigate whether trends are similar across Health Boards or if changes arise due to some parts of Scotland being colder and dimmer compared to others.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Data packages used
library(tidyverse)
library(janitor)
library(gt)
library(here)
```

```{r}
#Open data - January 2025 to June 2025 
data_jan_to_june2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

#Health Boards names and codes
HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()
```
Keep our focus on SSRIs of all formulations
```{r}
#filter data 
data_jan_to_june2025 <- data_jan_to_june2025 %>% 
  filter(str_detect(bnf_item_description, "SERTRALINE|CITALOPRAM|ESCITALOPRAM|FLUOXETINE|PAROXETINE")) %>%  full_join(HB_lookup, by = c("hbt" = "hb")) %>% 
   select(hb_name, hbt:paid_date_month) 

```
Here I will be removing the last row
```{r}
#total per HB
total_AD_per_HB <- data_jan_to_june2025 %>% 
  group_by(hb_name) %>% 
  summarise(total_per_HB = sum(paid_quantity, na.rm = TRUE)) %>% 
  gt() %>% 
  cols_label(total_per_HB = "Total AD quantity",
             hb_name = "Health Board") %>% 
  print(total_AD_per_HB)
```
Here I will be removing the last row and change the row names in the 'Months' column
```{r}
#group by paid month date
total_AD_per_month <- data_jan_to_june2025 %>% 
  group_by(paid_date_month) %>% 
  summarise(total_per_month = sum(paid_quantity, na.rm = TRUE)) %>% 
  gt() %>% 
  cols_label(total_per_month = "Total AD quantity",
             paid_date_month = "Months") %>% 
  print(total_AD_per_month)
```

I'm thinking of including both graphs and maps
To account for the different population numbers in the different Health Boards, I'm thinking of joining the population data from the census to calculate the estimate AD prescription per person using the mean of the population from all the Health Boards.