---
title: "Do anti-depressants prescription rise in winter months and/or in the dimest
  regions of Scotland?"
author: "B209732"
date: "2025-11-14"
output:
  html_document: default
  pdf_document: default
---

Seasonal affective disorder (SAD) is a type of depression also known as the 'winter depression' because it's symptoms are more prevalent and severe during winter months. According to the NHS website, one treatment for SAD is the prescription of anti-depressants (ADs) such as selective serotonin re-uptake inhibitors (SSRIs) (https://www.nhs.uk/mental-health/conditions/seasonal-affective-disorder-sad/overview/ - accessed November 13th, 2025)

This report is going to investigate the link between AD prescriptions from January to June 2025, to see whether prescriptions were higher during winter months (January and February) compared to spring (March to June). Prescription data will be subdivided into Health Boards to investigate whether trends are similar across Health Boards or if changes arise due to some parts of Scotland being colder and dimmer compared to others.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Data packages used
library(tidyverse)
library(janitor)
library(gt)
library(here)
```

```{r}
#Open data - January 2025 to June 2025 
data_jan_to_june2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

#Health Boards names and codes
HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()
```
Keep our focus on SSRIs of all formulations
```{r}
#filter data 
data_jan_to_june2025 <- data_jan_to_june2025 %>% 
  filter(str_detect(bnf_item_description, "SERTRALINE|CITALOPRAM|ESCITALOPRAM|FLUOXETINE|PAROXETINE")) %>%  full_join(HB_lookup, by = c("hbt" = "hb")) %>% 
   select(hb_name, paid_quantity, paid_date_month)
  

```

```{r}
#total per HB
data_jan_to_june2025 %>% 
  select(hb_name, paid_quantity) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  ggplot(aes(x = hb_name, y = paid_quantity)) +
  geom_col() +
  labs(y = "Paid quantity", 
       x = "Health Boards",
       title = "Anti-depressants prescription in each Health Board") +
  theme_light()+
  coord_flip()

```
This graph shows that the greatest number of AD prescription is attributed to NHS Greater Glasgow and Clyde followed by NHS Lothian. This is probably because these Health Boards have higher populations compared to some other Health Boards. 

Next steps:
- to show differences in population, the population data will be plotted
- then the population data will be joined to the main dataset and AD prescription per person using the mean population across all Health Boards will be calculated and plotted - this shows whether colder regions of Scotland or those regions with less hours of daylight have higher AD prescriptions.
- the main dataset will then be used to plot a graph showing differences in AD prescriptions across months

- I'm also hoping to use maps
- I may use tables to aid comparisons of numbers

