---
title: "Do anti-depressants prescription rise in winter months and/or in the dimest
  regions of Scotland?"
author: "B209732"
date: "2025-11-14"
output:
  pdf_document: default
  html_document: default
---

Seasonal affective disorder (SAD) is a type of depression also known as the 'winter depression' because it's symptoms are more prevalent and severe during winter months. According to the NHS website, one treatment for SAD is the prescription of anti-depressants (ADs) such as selective serotonin re-uptake inhibitors (SSRIs) (https://www.nhs.uk/mental-health/conditions/seasonal-affective-disorder-sad/overview/ - accessed November 13th, 2025)

This report is going to investigate the link between AD prescriptions from January to June 2025, to see whether prescriptions were higher during winter months (January and February) compared to spring (March to June). Prescription data will be subdivided into Health Boards to investigate whether trends are similar across Health Boards or if changes arise due to some parts of Scotland being colder and dimmer compared to others.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Data packages used
library(tidyverse)
library(janitor)
library(gt)
library(here)
```

```{r}
#Open data - January 2025 to June 2025 
data_jan_to_june2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

#Health Boards names and codes
HB_lookup <- read_csv("https://www.opendata.nhs.scot/dataset/9f942fdb-e59e-44f5-b534-d6e17229cc7b/resource/652ff726-e676-4a20-abda-435b98dd7bdc/download/hb14_hb19.csv") %>% 
  clean_names()
```
Keep our focus on SSRIs of all formulations
```{r}
#filter data 
data_jan_to_june2025 <- data_jan_to_june2025 %>% 
  filter(str_detect(bnf_item_description, "SERTRALINE|CITALOPRAM|ESCITALOPRAM|FLUOXETINE|PAROXETINE")) %>%  full_join(HB_lookup, by = c("hbt" = "hb")) %>% 
   select(hb_name, paid_quantity, paid_date_month)
  

```

```{r}
#total per HB
data_jan_to_june2025 %>% 
  select(hb_name, paid_quantity) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  summarise(total_per_HB = sum(paid_quantity, na.rm = TRUE)) %>% 
    gt() %>% 
  cols_label(total_per_HB = "Total AD quantity",
             hb_name = "Health Board") %>% 
  tab_header(title = "Total anti-depressants prescrption",
             subtitle = "Data from NHS Borders and NHS Fife")

data_jan_to_june2025 %>% 
  select(hb_name, paid_quantity) %>% 
  drop_na()%>% #to remove missing values
  group_by(hb_name) %>% 
  ggplot(aes(x = hb_name, y = paid_quantity)) +
  geom_col() +
  labs(y = "Paid quantity", 
       x = "Health Boards",
       title = "Anti-depressants prescription in each Health Board") +
  theme_light()+
  coord_flip()

```

I'm thinking of including both graphs and maps
To show that the differences in paid quantity could be due becase of differences in population, I'm going to plot the population data for the different Health Boards.
To account for the different population numbers in the different Health Boards, I'm thinking of joining the population data from the census to calculate the estimate AD prescription per person using the mean of the population from all the Health Boards.